---
title: "Quarto: a crash course"
subtitle: "Writing reproducible, branded reports (with or without Quarto and usedepartmenttheme)"
author: "Jack Goodall"
date: last-modified
date-format: "[Last rendered] DD-MM-YYYY"
bibliography: references.bib
link-citations: true
format:
  html:
    toc: true
    toc-depth: 4
    code-fold: true
    code-tools: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
# Options that apply to the whole document.
options(scipen = 999, digits = 2)  # no scientific notation; round to 2 dp
```

# Overview

This is a crash course in [Quarto](https://quarto.org) — the next-generation,
multi-language successor to R Markdown — and a guide to applying your
department's branding with the
[**usedepartmenttheme**](https://github.com/jackwgoodall/usedepartmenttheme)
package.

Although it is aimed at users of `usedepartmenttheme`, I hope it is of broader
interest, and anyone is welcome to use and adapt it. It has three aims:

1.  To show what a `usedepartmenttheme` document looks like when rendered.
2.  To explain how to install, use, and customise the themes.
3.  To provide a tutorial on Quarto for people who know R but are new to
    Quarto (or coming from R Markdown).

If you are new to markdown, I also hope to convert you: Quarto is a
powerful way to communicate pre-publication research findings with colleagues
and collaborators, wrapping a narrative around your data and code.

If you are already comfortable with R Markdown or Quarto, the
[Applying a theme](#applying-a-department-theme) section is all you need to get
going. If you are new to it, I suggest reading through in order.
If you don't care about the `usedepartmenttheme` you can skip to [Anatomy of a Quarto document](#anatomy). 

For more depth (including how to make PDFs, Word documents, books, websites,
or dashboards) the [Quarto guide](https://quarto.org/docs/guide/) is excellent,
as is the [R Markdown Cookbook](https://bookdown.org/yihui/rmarkdown-cookbook/),
most of which still applies.

In homage to the original R Markdown tutorial, the examples use base R's
built-in `pressure` and `cars` datasets [@mcneil1977interactive].

If you have questions, suggestions, or would like to contribute, please
[open an issue](https://github.com/jackwgoodall/usedepartmenttheme/issues) on
the GitHub page.

::: {.callout-note}
## Quarto or R Markdown?

Quarto is the successor to R Markdown. Documents use the `.qmd` extension and
most of your R Markdown knowledge carries straight over, with a few syntax
refreshes noted throughout this page (chunk options move to `#|` comments,
inline code gains braces, tabsets become fenced `::: {.panel-tabset}`
blocks, and `output:` becomes `format:`). Unlike R Markdown, one Quarto install
works across R, Python, and Julia, and ships its own command-line tool, so it
does not depend on any single R package.
:::

# Applying a department theme {#applying-a-department-theme}

`usedepartmenttheme` installs a branded Quarto HTML extension into your project
with a single function call. Your `.qmd` files then need only a `format:` key.

1.  Install the package (installs `pak` first if you don't have it):

    ```{r}
    #| eval: false
    if (!require("pak", quietly = TRUE)) install.packages("pak")
    pak::pkg_install("jackwgoodall/usedepartmenttheme")
    ```

2.  In your **project root** (the folder containing your `.qmd` files), run the
    installer for your department once. For the Florey Institute:

    ```{r}
    #| eval: false
    usedepartmenttheme::florey()
    ```

    Other themes are `lshtm()`, `lshtm_mrcg()` (MRC Unit The Gambia), and
    `lshtm_mrcu()` (MRC Unit Uganda). This creates an `_extensions/` folder
    (commit it to version control so collaborators can render without the
    package installed) and a `_quarto.yml`.

3.  Add the matching `format:` key to your document's YAML (the Console output provides you with this):

    ``` yaml
    ---
    title: "My analysis"
    format: florey-html
    ---
    ```

4.  Render (the **Render** button in RStudio, or `quarto render`). You should
    get an HTML file with the department's branded header bar, logo, and a
    matching browser-tab favicon.

## Overriding the colour, logo, or favicon

Every theme accepts a `colour`, `logo`, and `favicon`:

```{r}
#| eval: false
usedepartmenttheme::florey(logo = "images/my-logo.png", colour = "#450099")
```

## Not one of the departments?

Use `custom()` with your own base colour and logo — the package derives the
lighter/paler tints and a readable header-text colour automatically:

```{r}
#| eval: false
usedepartmenttheme::custom(colour = "#003865", logo = "images/logo.png",
                           name = "My Lab")
```

::: {.callout-tip}
The themes impose no document defaults of their own — options like `toc`,
`code_fold`, and `self_contained` are opt-in arguments on every installer.
See the [package README](https://github.com/jackwgoodall/usedepartmenttheme)
for the full list.
:::

# Anatomy of a Quarto document {#anatomy}

A Quarto document is made up of:

-   the YAML header,
-   formatted text,
-   code chunks.

The **YAML header** (*"YAML Ain't Markup Language"* — recursive acronyms are
big in coding circles) tells Quarto how to render the document. The only
essential entry is `format:`; I'd suggest `html` for almost all tasks (though
PDF and Word have their uses). There is much more the header can do to make
documents more accessible, as you'll see.

**Formatted text** is what makes Quarto so good for sharing pre-publication
work: it lets you build a narrative around the data you want to present. The
options are shown in the [Formatting](#formatting) section.

**Code chunks** are sections of R code that either set things up or produce an
output to show. Code inside a chunk behaves much like normal R — but plots and
console output appear directly after the chunk that created them.

# Code chunks {#code-chunks}

## Inserting a chunk

All executable code lives in code chunks, framed by ```` ```{r} ```` at the
start and ```` ``` ```` at the end. Insert one manually, with the **Insert**
menu, or with `Ctrl`/`Cmd` + `Alt` + `I`.

You can give a chunk a label after the `r`, which makes it easier to find later
(it appears in the editor's outline). The chunk below shows its own source (via
the `echo: fenced` option) followed by its result:

```{r}
#| label: chunk-one
#| echo: fenced
1 + 1
```

## Setting chunk options

In Quarto, chunk options are written *inside* the chunk on special `#|`
("hash-pipe") comment lines, rather than in the chunk header as in R Markdown.
They sit just below the opening ```` ```{r} ````. For example, these lines label
a chunk and hide its code while keeping the output:

``` r
#| label: chunk-two
#| echo: false
```

Rather than repeat options on every chunk, set document-wide defaults in the
YAML `execute:` block (see the top of this file):

``` yaml
execute:
  warning: false
  message: false
```

This document uses:

-   `code-fold: true` in the YAML, so code is shown but hidden behind a
    **Code** toggle by default — a nice compromise between collaborators who
    want to see the code and those who just want the content. Click **▸ Code**
    on any chunk to expand it.
-   `warning: false` and `message: false`, so warnings and messages are hidden
    in the output. Watch for them while you're *building* the document, as they
    won't surface in the render.
-   `options(scipen = 999, digits = 2)` (in the hidden `setup` chunk) to switch
    off scientific notation and round to two decimal places. Adjust to taste.

The most useful per-chunk options are:

| Option | Effect |
|---|---|
| `echo: false` | Hide the code, keep its output |
| `eval: false` | Show the code but don't run it |
| `include: false` | Run the code but hide both it and its output |
| `output: false` | Run and show the code but hide its output |
| `warning: false` / `message: false` | Hide warnings / messages |

A per-chunk option always takes precedence over the document default. So to
show every chunk except one, keep the default and add `#| echo: false` to that
one chunk.

If you'll be doing heavy data processing, it's also worth reading about
[caching](https://quarto.org/docs/computations/caching.html) and Quarto's
[freeze](https://quarto.org/docs/projects/code-execution.html#freeze) feature.

## Loading packages

Packages used in the render must be loaded *in the document*, not just in your
session. A robust pattern is to bootstrap `pacman` and let it install and load
everything else:

```{r}
#| label: packages
if (!require("pacman", quietly = TRUE)) install.packages("pacman")
pacman::p_load(tidyverse, kableExtra)
```

I also set a global `ggplot2` theme once, up front, so every plot inherits it
(discussed in [Plot themes](#plot-themes)):

```{r}
#| label: ggplot-theme
theme_set(
  theme_minimal() +
    theme(
      panel.grid.major = element_blank(),        # remove major grid lines
      panel.grid.minor = element_blank(),        # remove minor grid lines
      panel.background = element_blank(),         # remove the background
      axis.line        = element_line(colour = "black")  # solid axis line
    )
)
```

# Formatting {#formatting}

Create headings with increasing numbers of hashes (`#`) — the more hashes, the
smaller the heading. Anything up to level 4 (`####`) appears in the floating
table of contents (controlled by `toc-depth:` in the YAML).

## Headings

This is heading 2 (`##`).

### Heading 3

This is heading 3 (`###`).

#### Heading 4

This is heading 4 (`####`).

##### Heading 5

Heading 5 (`#####`) still renders, but won't appear in the table of contents
given our `toc-depth: 4`.

## Text options

| Input | Output |
|---|:---:|
| `**Two asterisks**` | **makes text bold** |
| `*One asterisk*` | *italicises text* |
| `***Three asterisks***` | ***does both*** |
| `` `Single backticks` `` | `format as code` |
| `~~Two tildes~~` | ~~cross out~~ |

Links are made by putting the text in square brackets followed by the URL in
round brackets, with no space between:

``` markdown
[Link](https://github.com/jackwgoodall/usedepartmenttheme/)
```

makes [Link](https://github.com/jackwgoodall/usedepartmenttheme/).

Equations are wrapped in `$` for inline and `$$` for displayed maths:

$$
\sum_{i = 1}^{n}{(this..is..an..equation)^2}
$$

A reference for mathematical notation is
[here](https://rpruim.github.io/s341/S19/from-class/MathinRmd.html).

## Dividing lines

Three or more asterisks make a horizontal rule:

***

A heavier divide can be made with a little HTML:

``` html
<hr style="border: 1px solid black; width: 100%; margin: 1em 0;">
```

<hr style="border: 1px solid black; width: 100%; margin: 1em 0;">

(swap `black` for another colour name or hex code to match your style).

## Spaces

Markdown largely ignores whitespace. Two spaces at the end of a line start a
new line; for something more explicit (or multiple blank lines) insert `<br>`.
`&nbsp;` inserts a non-breaking space.

## Tables

Tables can be written by hand (as in [Text options](#text-options) above), but
more elegant ones come from piping a data frame into **kableExtra**:

```{r}
#| label: kable-demo
table(pressure$temperature[1:5], pressure$pressure[1:5]) |>
  kbl(caption = "A hover table") |>
  kable_paper("hover", full_width = FALSE)
```

Many more options are documented on the
[kableExtra site](https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html).

# R code in formatted text

You can drop values from R objects straight into your prose, so statements
update automatically when the data changes. Inline code in Quarto is written in
back ticks with a leading `{r}` marker:

``` markdown
The median pressure is `{r} median(pressure$pressure)`.
```

which renders as:

> The median pressure of the `pressure` dataset is
> `{r} median(pressure$pressure)`.

If the data changed, that number would update on the next render. (This is a
change from R Markdown, where the marker was just `r`, without the braces.)

# Tabsets

Tabs are one of the most effective ways to make a data-rich document
approachable. In Quarto, wrap a set of same-level headings in a fenced
`::: {.panel-tabset}` block; each heading becomes a tab:

::: {.panel-tabset}

## Standard

```{r}
#| label: pressure-standard
p <- pressure |>
  ggplot(aes(x = temperature, y = pressure)) +
  geom_point()
p
```

## Red dots

```{r}
#| label: pressure-red
p + geom_point(colour = "red", size = 4)
```

## Blue triangles

```{r}
#| label: pressure-blue
p + geom_point(colour = "#093daf", size = 4, shape = 17)
```

:::

(In R Markdown this was done with a `{.tabset}` attribute on the parent
heading; the fenced-div approach is the Quarto equivalent.)

# Plots

## Absolute plot size

Set the rendered size with the `fig-width` and `fig-height` chunk options (in
inches):

```{r}
#| label: fig-abs
#| fig-width: 2
#| fig-height: 3
p + geom_point(colour = "#093daf", size = 4, shape = 17)
```

## Relative plot size

How much of the page a figure occupies is set with `out-width` (and
`out-height`):

```{r}
#| label: fig-rel
#| out-width: "50%"
p + geom_point(colour = "#093daf", size = 4, shape = 17)
```

## Side by side

Quarto lays multiple outputs out for you with `layout-ncol`:

```{r}
#| label: fig-side-by-side
#| layout-ncol: 2
#| fig-cap:
#|   - "Figure 1"
#|   - "Figure 2"
p + geom_point(colour = "#093daf", size = 4, shape = 17)
p + geom_point(colour = "#c21033", size = 4, shape = 17)
```

(This replaces R Markdown's `out.width = "50%"` + `fig.show = "hold"` recipe.)

## Plot themes {#plot-themes}

As in ordinary ggplot2 code you can switch themes (`theme_minimal()` is
popular). Because we called `theme_set()` at the top of the document, every
plot inherits that look unless it specifies its own:

```{r}
#| label: fig-themes
#| layout-ncol: 2
p + geom_point(colour = "#c21033", size = 4, shape = 17) +
  ggtitle("This…") +
  theme(plot.title = element_text(size = 24))

p + geom_point(colour = "#c21033", size = 4, shape = 17) +
  ggtitle("…rather than this") +
  theme_grey() +
  theme(plot.title = element_text(size = 24))
```

# Callouts

Quarto's callout blocks are a clean way to flag notes, tips, and warnings — a
feature R Markdown lacks. Write them as fenced divs:

``` markdown
::: {.callout-warning}
Mind the gap.
:::
```

::: {.callout-note}
A neutral note.
:::

::: {.callout-tip}
A helpful tip.
:::

::: {.callout-warning}
Something to be careful about.
:::

# References

## Style

Citations come from a BibTeX file referenced in the YAML (`bibliography:`). The
formatting style can be changed with a Citation Style Language file: download
one from [Zotero's style repository](https://www.zotero.org/styles), put it in
your project, and point `csl:` at it. For example, to use Nature's style:

``` yaml
bibliography: references.bib
csl: nature.csl
link-citations: true
```

## Building the BibTeX file

The easiest way to build a `.bib` file is a reference manager such as Zotero —
right-click one or more items to export, or **File → Export Library** for the
whole thing. Online tools can also turn DOIs into BibTeX.

## Citing in the text

An inline citation is an `@` followed by the reference key (`mcneil1977interactive`
here), which renders as @mcneil1977interactive. Because we set
`link-citations: true`, it hyperlinks to the bibliography. Adding a
`# References` heading at the end of the document renders the bibliography there
automatically.

# Cross-references

Link to another section by putting text in square brackets followed by the
heading's slug — lower case, no punctuation, spaces as hyphens:

``` markdown
[the formatting section](#formatting)
```

gives [the formatting section](#formatting). Quarto also has a richer
cross-referencing system (label a section `{#sec-foo}` and refer to it with
`@sec-foo` when `number-sections: true`); see the
[Quarto docs](https://quarto.org/docs/authoring/cross-references.html).

# The YAML, annotated

Most of these have appeared already; here is the full header for this document
with a note on each line. YAML is space-sensitive — child lines are indented
under their parent.

``` yaml
title: "Quarto: a crash course"        # document title
subtitle: "…"                          # optional subtitle
author: "Jack Goodall"                 # your name
date: last-modified                    # auto-fills the last render date
date-format: "[Last rendered] DD-MM-YYYY"  # UK-style date
bibliography: references.bib           # BibTeX source for citations
link-citations: true                   # hyperlink citations to the bibliography
format:
  html:                                # render to HTML
    toc: true                          # add a table of contents
    toc-depth: 4                       # include headings down to level 4
    code-fold: true                    # show code, but folded by default
    code-tools: true                   # add the document code menu
execute:
  warning: false                       # hide warnings document-wide
  message: false                       # hide messages document-wide
```

When you install a `usedepartmenttheme` extension, you'd swap the whole
`format:` block for a single key such as `format: florey-html`, and the theme
supplies the styling, header, and favicon for you.

# Conclusions

I hope this crash course has been useful. I'd love your feedback — good or bad —
via an [issue on the package page](https://github.com/jackwgoodall/usedepartmenttheme/issues).

# References

::: {#refs}
:::
